---
title: "dplyr and Arrow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr and Arrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("arrow", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("geoarrow", quietly = TRUE) &&
  isTRUE(arrow::arrow_info()$capabilities[["acero"]])
knitr::opts_chunk$set(eval = can_run)
```

Spatial functions that run inside Arrow's engine, not in R.

**Why it matters:** Arrow only knows the functions registered with it. Call
anything else on a `Table` and it warns, then pulls your data into R. On a
`Dataset` bigger than memory, that fails.

```{r}
#| message: false
library(geoarrowrs)
library(arrow)
library(dplyr)
library(geoarrow)
```

## Register once

```{r}
nc <- as.data.frame(read_shapefile(
  system.file("shape/nc.shp", package = "sf")
))
tbl <- arrow_table(nc)

register_geoarrow_udfs(tbl)
```

## Then write normal dplyr

```{r}
tbl |>
  mutate(area = unsigned_area(geometry)) |>
  select(NAME, area) |>
  head(3) |>
  collect()
```

**Filter on a predicate.** The comparison happens in the engine.

```{r}
tbl |>
  filter(unsigned_area(geometry) > 0.1) |>
  summarise(n = n()) |>
  collect()
```

**Geometry in, geometry out.** The result keeps its GeoArrow type.

```{r}
tbl |>
  mutate(centre = centroid(geometry)) |>
  pull(centre) |>
  class()
```

**Datasets stream.** Nothing here holds the column in memory.

```{r}
dir <- file.path(tempdir(), "nc-dataset")
unlink(dir, recursive = TRUE)
write_dataset(tbl, dir, format = "feather")

open_dataset(dir, format = "feather") |>
  filter(unsigned_area(geometry) > 0.1) |>
  summarise(n = n()) |>
  collect()
```

## What gets skipped, and why

A warning names anything that could not register. Three reasons:

- **Wrong type.** `length_euclidean()` needs linestrings. Against
  multipolygons there is nothing to measure.
- **Non-geometry arguments.** `buffer()` takes a distance. These kernels take
  geometry columns only.
- **Aggregates.** `unary_union()` collapses an array to one row, which is not
  a scalar kernel.

## Two rules

**Register per geometry type.** Arrow dispatches on the exact input type, so
reading a different type means calling `register_geoarrow_udfs()` again with
that data.

**Use `prefix` to avoid collisions** with an existing Arrow binding.

```{r}
head(register_geoarrow_udfs(tbl, prefix = "geo_"), 3)
```

## The bottom line

One call, then ordinary dplyr. The geometry stays in Arrow.
