---
title: "dplyr and Arrow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dplyr and Arrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("arrow", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("geoarrow", quietly = TRUE) &&
  isTRUE(arrow::arrow_info()$capabilities[["acero"]])
knitr::opts_chunk$set(eval = can_run)
```

Spatial functions that run inside Arrow's engine, not in R.

**Why it matters:** Arrow only knows the functions registered with it. Call
anything else on a `Table` and it warns, then pulls your data into R. On a
`Dataset` bigger than memory, that fails.

```{r}
#| message: false
library(geoarrowrs)
library(arrow)
library(dplyr)
library(geoarrow)
```

## Register once

One call. Every function, every geometry type.

```{r}
nc <- as.data.frame(read_shapefile(
  system.file("shape/nc.shp", package = "sf")
))
tbl <- arrow_table(nc)

length(register_geoarrow_udfs(crs = tbl))
```

## Then write normal dplyr

```{r}
tbl |>
  mutate(area = unsigned_area(geometry)) |>
  select(NAME, area) |>
  head(3) |>
  collect()
```

**Filter on a predicate.** The comparison happens in the engine.

```{r}
tbl |>
  filter(unsigned_area(geometry) > 0.1) |>
  summarise(n = n()) |>
  collect()
```

**Geometry in, geometry out.** The result keeps its GeoArrow type and CRS.

```{r}
tbl |>
  mutate(centre = centroid(geometry)) |>
  pull(centre, as_vector = TRUE) |>
  class()
```

**Datasets stream.** Nothing here holds the column in memory.

```{r}
dir <- file.path(tempdir(), "nc-dataset")
unlink(dir, recursive = TRUE)
write_dataset(tbl, dir, format = "feather")

open_dataset(dir, format = "feather") |>
  filter(unsigned_area(geometry) > 0.1) |>
  summarise(n = n()) |>
  collect()
```

## Arguments that are not geometry

Numeric arguments take a literal or a column, and recycle row by row.

```{r}
tbl |>
  mutate(small = simplify(geometry, 0.05)) |>
  select(NAME, small) |>
  head(2) |>
  collect()
```

Option arguments take a literal. Arrow has no optional arguments, so give
every one.

```{r}
tbl |>
  mutate(dense = densify(geometry, 5000, "haversine")) |>
  summarise(pts = sum(n_coords(dense))) |>
  collect()
```

## The one thing to watch

**A GeoArrow type carries its CRS**, and Arrow matches kernels on the exact
type. Register for every CRS in one call, because registering a name again
replaces the kernels it had.

```{r}
plain <- arrow_table(
  id = 1:2,
  geometry = as_geoarrow_vctr(as_geoarrow_array(
    sf::st_sfc(sf::st_point(c(0, 0)), sf::st_point(c(1, 1)))
  ))
)

register_geoarrow_udfs(crs = list(NULL, tbl), functions = "unsigned_area")

bind_rows(
  collect(summarise(tbl, total = sum(unsigned_area(geometry)))),
  collect(summarise(plain, total = sum(unsigned_area(geometry))))
)
```

Call a function against a type it cannot handle and Arrow says so, rather
than silently falling back to R.

```{r}
#| error: true
tbl |>
  mutate(len = length_euclidean(geometry)) |>
  collect()
```

## Not registered

Three functions change the length of the array, which a scalar kernel may
not: `unary_union()`, `explode()`, and `flatten()`. Call those in R.

## The bottom line

One call, then ordinary dplyr. The geometry stays in Arrow.
