---
title: "Compute inside Arrow with dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compute inside Arrow with dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("arrow", quietly = TRUE) &&
  requireNamespace("dplyr", quietly = TRUE) &&
  requireNamespace("geoarrow", quietly = TRUE) &&
  isTRUE(arrow::arrow_info()$capabilities[["acero"]])
knitr::opts_chunk$set(eval = can_run)
```

Spatial functions that run in Arrow's query engine, not in R.

**Why it matters:** a `Dataset` can be larger than memory. If every spatial
call drags the geometry column into R, that advantage is gone.

```{r}
#| message: false
library(geoarrowrs)
library(arrow)
library(dplyr)
library(geoarrow)
```

## The problem

Arrow only knows the functions registered with it. Call anything else and it
gives up and pulls your data into R.

```{r}
nc <- as.data.frame(read_shapefile(
  system.file("shape/nc.shp", package = "sf")
))
tbl <- arrow_table(nc)
```

Without registration you get a warning and a fallback:

```
Warning: Expression not supported in Arrow
→ Pulling data into R
```

On a `Table` that is merely slow. On a `Dataset` that does not fit in memory,
it fails outright.

## The fix

One call.

```{r}
registered <- register_geoarrow_udfs(tbl)
length(registered)
```

Now the same code runs inside the engine.

```{r}
tbl |>
  mutate(area = unsigned_area(geometry)) |>
  select(NAME, area) |>
  head(3) |>
  collect()
```

## It composes

**Filter on a spatial predicate.** The comparison happens in the engine.

```{r}
tbl |>
  filter(unsigned_area(geometry) > 0.1) |>
  summarise(n = n()) |>
  collect()
```

**Geometry in, geometry out.** The result keeps its GeoArrow type.

```{r}
res <- tbl |>
  mutate(centre = centroid(geometry)) |>
  select(NAME, centre) |>
  head(3) |>
  collect()

class(res$centre)
```

**Predicates take two geometry columns.**

```{r}
tbl |>
  mutate(self = intersects(geometry, geometry)) |>
  summarise(all_true = all(self)) |>
  collect()
```

## Datasets stream

This is the case that matters. Nothing here holds the whole column in memory.

```{r}
dir <- file.path(tempdir(), "nc-dataset")
unlink(dir, recursive = TRUE)
write_dataset(tbl, dir, format = "feather")

open_dataset(dir, format = "feather") |>
  filter(unsigned_area(geometry) > 0.1) |>
  summarise(n = n()) |>
  collect()
```

## What gets registered

**46 of 51** functions register against a multipolygon column. The rest are
skipped on purpose, and the warning tells you which:

- **Wrong geometry type.** `length_euclidean()` needs linestrings. Against
  multipolygons there is nothing to measure, so it is not registered.
- **Non-geometry arguments.** `buffer()` and `simplify()` take a distance.
  Arrow kernels here only take geometry columns.
- **Aggregates.** `unary_union()` collapses an array to one row. That is not a
  scalar kernel.

Arrow dispatches on the exact input type, so a kernel is registered for the
precise GeoArrow type in your data. Reading a different geometry type means
calling `register_geoarrow_udfs()` again with that data.

Use `prefix` if a name would collide with an existing Arrow binding:

```{r}
head(register_geoarrow_udfs(tbl, prefix = "geo_"), 3)
```

## The bottom line

Register once, then write ordinary dplyr. The geometry stays in Arrow and the
work happens in the engine.
