---
title: "How fast is it?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How fast is it?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("geoarrow", quietly = TRUE) &&
  requireNamespace("nanoarrow", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

Algorithms are faster than `{sf}`. Readers are slower. Here are the numbers.

**Why it matters:** knowing which half is which tells you when this package
helps and when it does not.

```{r}
#| message: false
library(geoarrowrs)
library(sf)
library(geoarrow)
```

Every number below is measured when this page is built, on 5,000 polygons.

```{r}
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
big <- do.call(rbind, replicate(50, nc, simplify = FALSE))
geom <- st_geometry(big)
arr <- as_geoarrow_array(geom)

nrow(big)
```

```{r}
best <- function(f, n = 5) min(replicate(n, system.time(f())[["elapsed"]]))

compare <- function(...) {
  ops <- list(...)
  out <- do.call(rbind, lapply(names(ops), function(nm) {
    a <- best(ops[[nm]][[1]])
    b <- best(ops[[nm]][[2]])
    data.frame(op = nm, geoarrowrs = a, sf = b, speedup = round(b / a, 1))
  }))
  out
}
```

## Algorithms

```{r}
cv <- function(x) as.vector(nanoarrow::convert_array(x))

compare(
  area = list(
    function() cv(unsigned_area(arr)),
    function() as.numeric(st_area(big))
  ),
  centroid = list(
    function() centroid(arr),
    function() st_centroid(geom)
  ),
  convex_hull = list(
    function() convex_hull(arr),
    function() st_convex_hull(geom)
  ),
  simplify = list(
    function() simplify(arr, 0.01),
    function() st_simplify(geom, dTolerance = 0.01)
  )
)
```

**The pattern.** Anything that walks coordinates and returns a number is
several times faster. `convex_hull()` is closer to a tie, because the work is
dominated by the hull itself rather than by iteration overhead.

## Readers

This is the honest part.

```{r}
shp <- file.path(tempdir(), "bench.shp")
suppressWarnings(st_write(big, shp, quiet = TRUE, delete_dsn = TRUE))

compare(
  shapefile = list(
    function() as.data.frame(read_shapefile(shp)),
    function() st_read(shp, quiet = TRUE)
  )
)
```

**`{sf}` wins, by roughly 4x.** GDAL's shapefile reader is mature and hard to
beat. The cost here is in the read itself, not in handing the result to R:
building the stream without materialising it takes just as long.

So read speed is not the reason to use these readers. Two other things are:

- **The output is already Arrow.** No conversion step, and no R-side list of
  geometries.
- **FlatGeobuf can skip data.** Pass a bounding box and the packed Hilbert
  R-tree means unwanted features are never decoded.

```{r}
fgb <- file.path(tempdir(), "bench.fgb")
suppressWarnings(st_write(big, fgb, quiet = TRUE, delete_dsn = TRUE))

full <- best(function() as.data.frame(read_flatgeobuf(fgb)))
boxed <- best(function() {
  as.data.frame(read_flatgeobuf(fgb, bbox = c(-79, 35, -78, 36)))
})

round(full / boxed, 1)
```

That ratio is the index doing its job.

## Caveats

Read these before quoting any number above.

- **One machine, one dataset.** North Carolina counties repeated 50 times.
  Your geometries have different vertex counts and different complexity.
- **Timings are a minimum of five runs.** That favours warm caches, which is
  the right choice for comparing algorithms and a generous one for readers.
- **Nothing is parallel yet.** Both packages are single threaded here. The
  Rust side has room for `rayon` that has not been taken.
- **`{sf}` does more.** It carries a CRS through every operation and validates
  geometries. Some of its extra time buys something this package does not
  offer yet.

## The bottom line

Use this package for the compute and for staying in Arrow. Keep using
`{sf}` or GDAL when the bottleneck is reading a file.
