---
title: "How fast is it?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How fast is it?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("geoarrow", quietly = TRUE) &&
  requireNamespace("nanoarrow", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

Faster than `{sf}` at everything measured here. Often by a lot.

**Why it matters:** the geometry never becomes an R object. There is no list of
`sfg` classes to allocate, walk, and garbage collect, so the work happens at
Rust speed on contiguous Arrow buffers.

```{r}
#| message: false
library(geoarrowrs)
library(sf)
library(geoarrow)
```

Every number below is measured when this page is built.

```{r}
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
big <- do.call(rbind, replicate(200, nc, simplify = FALSE))
geom <- st_geometry(big)
arr <- as_geoarrow_array(geom)

format(nrow(big), big.mark = ",")
```

## Measuring honestly

Several of these operations finish in well under a millisecond, which
`system.time()` cannot resolve. Timing a single call would report `0.000` and
an infinite speedup.

So each expression is run enough times to take at least a quarter second, and
the total is divided by the count.

```{r}
per_call <- function(f, min_time = 0.25) {
  f()
  reps <- 1L
  repeat {
    elapsed <- system.time(for (i in seq_len(reps)) f())[["elapsed"]]
    if (elapsed >= min_time || reps >= 1e6L) break
    reps <- reps * 4L
  }
  elapsed / reps
}

compare <- function(...) {
  ops <- list(...)
  do.call(rbind, lapply(names(ops), function(nm) {
    a <- per_call(ops[[nm]][[1]])
    b <- per_call(ops[[nm]][[2]])
    data.frame(
      op = nm,
      geoarrowrs = signif(a, 3),
      sf = signif(b, 3),
      speedup = round(b / a, 1)
    )
  }))
}
```

## Algorithms

```{r}
cv <- function(x) as.vector(nanoarrow::convert_array(x))

compare(
  area = list(
    function() cv(unsigned_area(arr)),
    function() as.numeric(st_area(big))
  ),
  centroid = list(
    function() centroid(arr),
    function() st_centroid(geom)
  ),
  convex_hull = list(
    function() convex_hull(arr),
    function() st_convex_hull(geom)
  ),
  simplify = list(
    function() simplify(arr, 0.01),
    function() st_simplify(geom, dTolerance = 0.01)
  ),
  bounding_rect = list(
    function() bounding_rect(arr),
    function() lapply(geom, st_bbox)
  )
)
```

**The pattern.** The gap widens with how much per-geometry bookkeeping R has to
do. Operations returning a plain number are the most lopsided, because `{sf}`
builds and returns an R vector while allocating nothing on our side beyond one
Arrow buffer.

## Readers

```{r}
shp <- file.path(tempdir(), "bench.shp")
fgb <- file.path(tempdir(), "bench.fgb")
suppressWarnings({
  st_write(big, shp, quiet = TRUE, delete_dsn = TRUE)
  st_write(big, fgb, quiet = TRUE, delete_dsn = TRUE)
})

compare(
  shapefile = list(
    function() as.data.frame(read_shapefile(shp)),
    function() st_read(shp, quiet = TRUE)
  ),
  flatgeobuf = list(
    function() as.data.frame(read_flatgeobuf(fgb)),
    function() st_read(fgb, quiet = TRUE)
  )
)
```

Reading is faster too, and the gap grows with the file. GDAL constructs an R
object per feature; these readers fill Arrow buffers and hand back a stream.

**FlatGeobuf can also skip data outright.** Pass a bounding box and the packed
Hilbert R-tree means features outside it are never decoded.

```{r}
full <- per_call(function() as.data.frame(read_flatgeobuf(fgb)))
boxed <- per_call(function() {
  as.data.frame(read_flatgeobuf(fgb, bbox = c(-79, 35, -78, 36)))
})

round(full / boxed, 1)
```

That ratio is the index doing its job, and it scales with how selective the box
is.

## Benchmark against a release build

**This matters more than anything else on this page.**

`devtools::load_all()` compiles the Rust in debug. Debug Rust is not slow by a
little, it is slow by an order of magnitude, and benchmarking it will tell you
this package loses to `{sf}` at everything.

Measure an installed build:

```r
devtools::install()
# restart R, then benchmark
```

An early draft of this article was written against a `load_all()` session and
reported `{sf}` reading files four times faster. Rebuilt against an installed
package, the same comparison reversed.

## Caveats

- **One machine, one dataset.** North Carolina counties repeated 200 times.
  Your geometries have different vertex counts and different complexity.
- **Nothing is parallel yet.** Both packages are single threaded here. The Rust
  side has room for `rayon` that has not been taken.
- **`{sf}` does more.** It carries a CRS through every operation and validates
  geometries. Some of its extra time buys something this package does not
  offer yet.
- **Z and M are dropped** by every algorithm here, because the bridge to the
  `geo` crate goes through a two dimensional geometry. The readers preserve
  them.

## The bottom line

Staying in Arrow is the whole trick. Nothing is converted to an R object, so
nothing pays for it.
