---
title: "Read spatial files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Read spatial files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
can_run <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("geoarrow", quietly = TRUE) &&
  requireNamespace("nanoarrow", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

Three readers. One shape of output.

**Why it matters:** every reader hands back a record batch stream, so the data
is Arrow from the first byte. Nothing is converted to an R list of geometries
on the way in.

```{r}
#| message: false
library(geoarrowrs)
library(geoarrow)
```

## Shapefiles

`read_shapefile()` reads the `.shp` and its `.dbf` attributes together.

```{r}
path <- system.file("shape/nc.shp", package = "sf")
nc <- as.data.frame(read_shapefile(path))

dim(nc)
head(nc[, c("NAME", "BIR74", "geometry")], 3)
```

The attribute columns come first, `geometry` last.

**Z and M work.** Both are detected from the coordinate data, not the file
header. The header records a `0` range when M is absent, which is
indistinguishable from a real measure of zero, so it cannot be trusted.

## GeoJSON

```{r}
gj <- tempfile(fileext = ".geojson")
sf::st_write(sf::st_read(path, quiet = TRUE), gj, quiet = TRUE)

as.data.frame(read_geojson(gj))[1:3, c("NAME", "geometry")]
```

Two things worth knowing:

- **Property order is preserved.** Columns come back in document order, not
  sorted.
- **Types widen across features.** A property that is an integer in one feature
  and a double in the next becomes a double column.

## FlatGeobuf

The only reader that truly streams, and the only one with a spatial index.

```{r}
fgb <- tempfile(fileext = ".fgb")
sf::st_write(sf::st_read(path, quiet = TRUE), fgb, quiet = TRUE)

nrow(as.data.frame(read_flatgeobuf(fgb)))
```

Pass a bounding box and the reader uses the file's packed Hilbert R-tree to
skip everything outside it. It never decodes the features you did not ask for.

```{r}
nrow(as.data.frame(read_flatgeobuf(fgb, bbox = c(-79, 35, -78, 36))))
```

## What about GeoParquet?

Use `{geoarrow}` and `{arrow}`. It is already handled well there, so this
package does not duplicate it.

## The bottom line

Read into Arrow, stay in Arrow. The next article shows how to compute without
ever leaving it.
