---
title: "duckdbspatial"
---


The wonderful [`{duckspatial}`](https://cidree.github.io/duckspatial/) R package—which delegates operations to [DuckDB](https://duckdb.org/) has published bench marks for operations.

This article compares `{geoarrowrs}` implementations to `{sf}` and `{duckspatial}` by recreating the benchmark but including geoarrow-rs.

See the original [duckspatial benchmark](https://cidree.github.io/duckspatial/articles/benchmark.html).

### The earth is a sphere note a plane

One of the issues the duckdbspatial benchmark—is that is calculates **incorrect** distances.
`{sf}` delegates to `{s2}` to have very accurate spherical distance calculations—this comes at the cost of speed.

Using planar distances for calculating the distance from Jakarta to New York calculates the distance _through_ the earth. 

This issu

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE)

# pkgdown renders an article from the directory it lives in
root <- if (file.exists("bench/setup.R")) "." else "../.."
last_run <- file.path(root, "bench", "duckspatial", "last-run")

timings <- read.csv(file.path(last_run, "timings.csv"))
machine <- read.csv(file.path(last_run, "machine.csv"))

ops <- c("connect", "join", "filter", "intersects", "dissolve", "distance")
files <- c(
  "Shared setup" = "bench/duckspatial/setup.R",
  stats::setNames(
    file.path("bench/duckspatial", paste0(ops, ".R")),
    tools::toTitleCase(ops)
  )
)

# connect is a fixed charge rather than a row of the comparison, so it is read
# out on its own instead of becoming two mostly empty columns
connect <- timings[timings$op == "connect", ]
measured <- timings[timings$op != "connect", ]

wide <- stats::reshape(
  measured[c("op", "n", "pkg", "seconds")],
  idvar = c("op", "n"),
  timevar = "pkg",
  direction = "wide"
)
names(wide) <- sub("^seconds\\.", "", names(wide))
wide <- wide[order(match(wide$op, ops), wide$n), ]
wide$n <- format(wide$n, scientific = FALSE, big.mark = ",", trim = TRUE)
```

The five operations of the
[duckspatial benchmark](https://cidree.github.io/duckspatial/articles/benchmark.html),
run by `bench/duckspatial/run-all.R` and read back here from
`bench/duckspatial/last-run/`. Seconds, so lower is better.

```{r}
#| fig-width: 11
#| fig-height: 7
#| fig-alt: "Five log-log panels, one per operation, of elapsed seconds against
#|   input size. geoarrowrs is the lowest line in every panel. sf is the
#|   highest. duckspatial sits between them and is flat from 100,000 to one
#|   million rows, where its fixed startup cost is most of what is measured."
#| fig-cap: "Elapsed seconds by input size. Both axes are log scaled."

# base graphics rather than ggplot2, which 4.0.3 cannot load against the
# development version of S7
pal <- c(
  "geoarrowrs" = "#0072B2",
  "geoarrowrs (geodesic)" = "#56B4E9",
  "sedonadb" = "#CC79A7",
  "duckspatial" = "#009E73",
  "sf" = "#D55E00"
)

titles <- c(
  join = "Point in polygon join",
  filter = "Filter by polygon",
  intersects = "Intersects predicate",
  dissolve = "Dissolve by group",
  distance = "Cross distance (n x n)"
)

si <- function(x) {
  ifelse(
    x >= 1e6,
    paste0(x / 1e6, "M"),
    ifelse(x >= 1e3, paste0(x / 1e3, "k"), as.character(x))
  )
}

panel <- function(op) {
  d <- measured[measured$op == op, ]
  plot(
    NA,
    xlim = range(d$n),
    ylim = range(d$seconds),
    log = "xy",
    xaxt = "n",
    yaxt = "n",
    xlab = "",
    ylab = "",
    main = titles[[op]],
    font.main = 1,
    bty = "n"
  )
  ticks <- axTicks(2)
  abline(h = ticks, col = "grey92")
  axis(1, unique(d$n), si(unique(d$n)), tick = FALSE, col.axis = "grey30")
  axis(
    2,
    ticks,
    format(ticks, trim = TRUE, scientific = FALSE, drop0trailing = TRUE),
    las = 1,
    tick = FALSE,
    col.axis = "grey30"
  )

  for (pkg in names(pal)) {
    s <- d[d$pkg == pkg, ]
    if (!nrow(s)) {
      next
    }
    s <- s[order(s$n), ]
    lines(s$n, s$seconds, col = pal[[pkg]], lwd = 2)
    points(s$n, s$seconds, col = pal[[pkg]], pch = 19, cex = 1.1)
  }
}

withr::with_par(
  list(
    mfrow = c(2, 3),
    mar = c(3.5, 4, 3, 1.5),
    oma = c(2, 2, 0, 0),
    col.axis = "grey30",
    fg = "grey60"
  ),
  {
    for (op in names(titles)) {
      panel(op)
    }

    plot.new()
    legend(
      "center",
      names(pal),
      col = pal,
      lwd = 2,
      pch = 19,
      bty = "n",
      cex = 1.15,
      text.col = "grey20"
    )

    mtext("input size", 1, outer = TRUE, line = 0.5, cex = 0.8, col = "grey30")
    mtext("seconds", 2, outer = TRUE, line = 0.5, cex = 0.8, col = "grey30")
  }
)
```

```{r}
knitr::kable(t(machine), col.names = "")
knitr::kable(wide, row.names = FALSE)
```

Each implementation is timed until its answer exists in whatever form that
package keeps answers, starting from the same `sf` objects. The three do not
all return the same shape, and `rows` in `timings.csv` says what each produced.

```{r}
#| results: asis

# the accordion of https://pkg.garrickadenbuie.com/quarto-panel-accordion,
# written out rather than filtered in: that extension steps aside unless quarto
# supplies bootstrap itself, and pkgdown renders with `theme: none` because the
# site template brings its own. Divs rather than raw html throughout, so the
# code blocks inside still reach pandoc and get highlighted.
accordion <- "duckspatial-accordion"
fence <- strrep("`", 3)

cat("\n:::::: {.accordion #", accordion, "}\n\n", sep = "")

for (i in seq_along(files)) {
  id <- paste0(accordion, "-", i)
  path <- files[[i]]

  cat("::::: {.accordion-item}\n\n")
  cat('<h3 class="accordion-header">\n')
  cat(
    '<button class="accordion-button collapsed" type="button" ',
    'data-bs-toggle="collapse" data-bs-target="#',
    id,
    '" ',
    'aria-expanded="false" aria-controls="',
    id,
    '">',
    names(files)[[i]],
    "</button>\n",
    sep = ""
  )
  cat("</h3>\n\n")

  cat(
    ":::: {#",
    id,
    " .accordion-collapse .collapse ",
    'data-bs-parent="#',
    accordion,
    '"}\n\n',
    sep = ""
  )
  cat("::: {.accordion-body}\n\n")
  cat(fence, '{.r filename="', path, '"}\n', sep = "")
  cat(readLines(file.path(root, path)), sep = "\n")
  cat("\n", fence, "\n\n", sep = "")
  cat(":::\n\n::::\n\n:::::\n\n")
}

cat("::::::\n")
```
